---
title: "Untitled"
author: "Haoran Zhu"
date: "4/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# load and display data
data <- read.csv('GoemMixture.csv')
data <- data
head(data)
```

Because there are 0s in the dataset, to prevent $p_k$ exceeding 1, I wrote this EM algorithm slightly different from my exam result.


## EM algorithm

### Step 1: Initialize variables 

Choose random values for $c_k^{(0)}$ and $p_k^{(0)}$


### Step 2: E-Step

Recompute the assignment weight matrix as 

$$a_{ik}^{(j+1)}:=\frac{c_k^{(j)} f(x_i|p_k^{(j)})}{\sum\limits_{l=1}^K c_l^{(j)} f(x_i|p_k^{(j)})}.$$


### Step 3: M-Step

Recompute the proportions $c_k$ and the parameter $p_k$ as

$$c_k^{(j+1)}:=\frac{\sum\limits_{i=1}^n a_{ik}^{(j)}}{n}$$


$$p_k^{(j+1)}:=\frac{1}{1+E(x)}=\frac{1}{1 + \frac{\sum\limits_{i=1}^n a_{ik}^{(j+1)} x_i}{\sum\limits_{i=1}^n a_{ik}^{(j+1)}}}$$

### Step 4: Iterate till meet some criteria

The E-Step and M-Step are repeated alternatingly until convergence criterion (e.g. threshold) satisfied.


```{r}
calc.aik <- function(x, ck, pk, K) {
    result <- matrix(0, nrow = nrow(x), ncol = K)
    for (i in 1:nrow(x)){
        result[i,] <- ck * dgeom(x[i,], pk) / sum(ck * dgeom(x[i,], pk))
    }
    
    return(result)
}

expectation.maximization <- function(x, ck, pk, K, threshold = 1e-10, max_iter = 1000){
    aik.temp <- matrix(0, nrow = nrow(x), ncol = K)
    for (i in 1:max_iter) {
        # E-step
        aik <- calc.aik(x, ck, pk, K)
        
        # M-step
        # update ck
        ck <- apply(aik, 2, mean)
        
        # update pk
        for (i in 1:K){
            Ex <- sum(aik[,i] * x) / sum(aik[,i])
            pk[i] <-  1 / (1 + Ex)
        }

        # check if stop
        if (norm(aik-aik.temp, type='1') < threshold){
            return(list(ck=ck, pk=pk))
            break
        }
        
        # reassign aik temp for next comparison
        aik.temp <- aik
    }
    
    return(list(ck=ck, pk=pk))
}
```


```{r}
# calculate clusters after 
get.cluster <- function(data, em) {
  aik <- matrix(0, nrow = nrow(data), ncol = K)
  for (i in 1:nrow(data)){
      aik[i,] <- em$ck * dgeom(data[i,], em$pk) / sum(em$ck * dgeom(data[i,], em$pk))
  }
  
  cluster <- rep(0, nrow(data))
  for (i in 1:nrow(data)){
      cluster[i] <- which(aik[i,] == max(aik[i,]))
  }
  return(cluster)
}
```


```{r}
set.seed(2)
K <- 3
iter <- 20
ck.collect <- matrix(0, nrow = iter, ncol = K)
pk.collect <- matrix(0, nrow = iter, ncol = K)
likelihood.collect <- rep(0, iter)

for (i in 1:iter){
    print(paste('iteration #', i, sep=''))
    # get random ck and pk for starting
    ck.temp <- runif(K, 0, 1)
    ck.temp <- ck.temp / sum(ck.temp)
    pk.temp <- runif(K, 0, 1)
    
    # calculate the optimization
    em <- expectation.maximization(data, ck.temp, pk.temp, K)
    
    # store ck and pk
    ck.collect[i,] <- em$ck
    pk.collect[i,] <- em$pk
    
    # calculate cluster
    cluster <- get.cluster(data, em)
    
    # calculate and store likelihood 
    likelihood <- 0
    for (j in 1:nrow(data)){
        log.likelihood <- log(em$ck[1] * em$pk[1] * (1-em$pk[1])^(data[j,]) + 
                                em$ck[2] * em$pk[2] * (1-em$pk[2])^(data[j,])  + 
                                em$ck[3] * em$pk[3] * (1-em$pk[3])^(data[j,]) )
        likelihood <- likelihood + log.likelihood
    }
    likelihood.collect[i] <- likelihood
}



ck <- unique(ck.collect[which(likelihood.collect == max(likelihood.collect)), ])
pk <- unique(pk.collect[which(likelihood.collect == max(likelihood.collect)), ])
ck
pk
```

# plots

```{r}
library(ggplot2)
ggplot(data) +
  xlab("X") + ylab("") + ggtitle("plot of x") + 
  geom_point(mapping = aes(x = X, y = 0))
```


```{r}
cluster <- get.cluster(data, list(ck=ck, pk=pk))
ggplot(data) +
  xlab("X") + ylab("") + ggtitle("cluster of x") + 
  scale_color_discrete(name = 'cluster', labels = c('cluster1', 'cluster2', 'cluster3')) +
  geom_point(mapping = aes(x = X, y = 0, col = ifelse(cluster==1, "red", ifelse(cluster==2, 'blue', 'green'))))
```

Let's do a zoom-in for the left side. (We have to zoom in to see 3 clusters!)

```{r}
ggplot(data) +
  xlab("X") + ylab("") + ggtitle("cluster of x") + xlim(0,50) +
  scale_color_discrete(name = 'cluster', labels = c('cluster1', 'cluster2', 'cluster3')) +
  geom_point(mapping = aes(x = X, y = 0, col = ifelse(cluster==1, "red", ifelse(cluster==2, 'blue', 'green'))))
```
